---
title: "Clinical"
output: html_document
date: "2024-12-16"
---


## Load libraries
```{r, results = FALSE, message = FALSE, warning = FALSE}

library("TCGAbiolinks")
library("SummarizedExperiment")
```


## Download Data
```{r, results = 'hide', message = FALSE, warning = FALSE}

data.clinical.GBM <- GDCquery_clinic(
  project = "TCGA-GBM", 
  type = "Clinical"
)

data.clinical.LGG <- GDCquery_clinic(
  project = "TCGA-LGG", 
  type = "Clinical"
)
```



The clinical data's from the two 'project' do not have the same information.
```{r}

ncol(data.clinical.GBM)
ncol(data.clinical.LGG)

length(intersect(colnames(data.clinical.GBM), colnames(data.clinical.LGG)))

# columns that "data.clinical.GBM" have but not "data.clinical.LGG"
setdiff(colnames(data.clinical.GBM), colnames(data.clinical.LGG))
```




Merging the two, keeping the columns in common, the columns of "data.clinical.LGG".
```{r}

common_columns <- intersect(colnames(data.clinical.GBM), colnames(data.clinical.LGG))
data.clinical <- rbind(data.clinical.GBM[, common_columns], data.clinical.LGG[, common_columns])

clinical <- as.matrix(data.clinical)
rownames(clinical) <- clinical[,"submitter_id"]
clinical <- clinical[,-2]

rm(data.clinical.LGG)
rm(data.clinical.GBM)
```
They contain survival data!


In clinical data, the samples identifier are indeed "submitter_id"
Note that in https://docs.gdc.cancer.gov/Data/Data_Model/GDC_Data_Model/, it says that "submitter_id" are unique for the pair "project_type" and "submitter_id" (pair PK).
```{r}

length(unique(data.clinical[,"submitter_id"])) == length(data.clinical[,"submitter_id"])
rm(data.clinical)
```
It seems we don't have that problem here.


```{r}

clinical <- as.data.frame(clinical)
```


## Explore

```{r}

colnames(clinical)

unique(clinical$classification_of_tumor)
unique(clinical$tumor_grade)
unique(clinical$primary_diagnosis) # is this the classification? or just rely on the project?
unique(clinical$tumor_grade)
```



Classification of Samples
```{r}
labels <- clinical$primary_diagnosis

transformed_labels <- ifelse(labels == "Glioblastoma", "glioblastoma",
                      ifelse(labels %in% c("Oligodendroglioma, anaplastic", "Oligodendroglioma, NOS"), "oligodendroglioma",
                      ifelse(labels %in% c("Astrocytoma, NOS", "Astrocytoma, anaplastic"), "astrocytoma",
                      ifelse(labels %in% c("Mixed glioma"), "mixed glioma",
                      labels))))

clinical <- cbind(clinical, transformed_labels)
```


```{r}

summary(as.factor(clinical$transformed_labels))
print("------")
summary(as.factor(clinical$primary_diagnosis))
```




- Is the tumor classification correct, I mean, associating with the right project?
```{r}

clinical$project[clinical$transformed_labels == "glioblastoma"]
clinical$project[clinical$transformed_labels == "astrocytoma"]
clinical$project[clinical$transformed_labels == "oligodendroglioma"]
clinical$project[clinical$transformed_labels == "mixed glioma"]
```
ANSWER: Yes, it is!


- What it means "Mixed glioma"? From a specific project? 

https://www.cancer.gov/publications/dictionaries/cancer-terms/def/mixed-glioma 
ANSWER: is a mix of astrocytoma and oligodendroglioma (but LGG cancer).


```{r}

important_columns <- c("project", "days_to_last_follow_up", "days_to_death", "age_at_index", "vital_status", "gender", "primary_diagnosis", "transformed_labels")
clinical <- clinical[,important_columns]
```


Replace Not Reported by NA (in "vital_status" and "gender") 
```{r}

clinical$gender[clinical$gender == "not reported"] <- NA
clinical$vital_status[clinical$vital_status == "Not Reported"] <- NA
```




# Survival Data
```{r}

createSurvivalAssay <- function(clinical_assay){
  
  assay.survival <- matrix(data = NA, nrow = nrow(clinical_assay), ncol = 2,
                            dimnames = list(rownames(clinical_assay), c("survival", "death")))
  
  assay.survival[, "death"] <- ifelse(
    clinical_assay[, "vital_status"] == "Dead", 1,
    ifelse(
      is.na(clinical_assay[, "vital_status"]) | clinical_assay[, "vital_status"] == "Not Reported",
      NA,
      0
    )
  )
  
  assay.survival[,"survival"] <- ifelse(
    assay.survival[,"death"] == 1, 
    clinical_assay[,"days_to_death"],
    ifelse(
      assay.survival[,"death"] == 0,
      clinical_assay[,"days_to_last_follow_up"],
      NA
    )
  )
  
  return(as.data.frame(assay.survival))
}

```

```{r}

survival <- createSurvivalAssay(clinical)
dim(survival)
```


```{r}
colnames(survival) <- c("Time", "Status")
```

```{r}

survival.complete <- survival[complete.cases(survival), ]
dim(survival.complete)
```

```{r}
write.csv(survival.complete, "DataSets/processed_assays/survival_complete.csv", row.names = TRUE)
```

## Save final matrix
```{r}

write.csv(clinical, "DataSets/processed_assays/clinical.csv", row.names = TRUE)
```

