---
title: "RNA"
output: html_document
date: "2024-11-26"
---


## Load libraries
```{r, results = FALSE, message = FALSE, warning = FALSE}

library("TCGAbiolinks")
library("SummarizedExperiment")

library("DESeq2")
library("edgeR")
```


## Load useful functions

```{r}

ids_duplicated <- function(assay){
  
  prefixes <- substr(rownames(assay), 1, 12)
  groups <- split(rownames(assay), prefixes)
  duplicated_groups <- groups[sapply(groups, length) > 1]
  return(duplicated_groups)
}

  
del_repl_barcodes <- function(assay) {
  
  dupl_barcodes <- unlist(ids_duplicated(assay), use.names = FALSE)
  filtered_assay <- assay[!rownames(assay) %in% dupl_barcodes, , drop = FALSE]
  rownames(filtered_assay) <- substr(rownames(filtered_assay), 1, 12)
  
  return (filtered_assay)
}

get_assay_gene_type <- function(genes_types, assay, info) {
  
  # find the indices of the genes with the matching 'genes_types'
  selected_genes <- which(info[,"gene_type"] %in% genes_types)
  
  # select the corresponding columns from the assay matrix
  selected_assay <- assay[, selected_genes, drop = FALSE]
  return(selected_assay)
}

```



## Download Data
There are (for 'data.category' = "Transcriptome Profiling"), the following 'data.type':
  . Differential Gene Expression (only 44 files)
  . Gene Expression Quantification (~24k): contains quantified expression levels of all genes in the sample (for broad analysis of gene expression).
  . Isoform Expression Quantification (~ 17k)
  . miRNA Expression Quantification (~ 17k): quantified expression levels of microRNAs, small non-coding RNAs that regulate gene expression.
  . Raw Intensities (~1k files)
  . Single Cell Analysis (only 88 files)
  
Setting 'data.type' to "Gene Expression Quantification". 
```{r, results = 'hide', message = FALSE, warning = FALSE}

query_rna <- GDCquery(
  project = c("TCGA-GBM", "TCGA-LGG"),
  data.category = "Transcriptome Profiling" ,
  data.type = "Gene Expression Quantification" ,
  workflow.type = "STAR - Counts", # the others 2 options very few files
  access = "open",
  platform = "Illumina", # only one available (for this 'data.type')
  data.format = "tsv", # only one available (for this 'workflow.type')
  experimental.strategy = "RNA-Seq" # only one available (for this 'workflow.type')
)

if (!dir.exists("DataSets/GDCdata/TCGA-GBM/Transcriptome_Profiling/Gene_Expression_Quantification") ||
    !dir.exists("DataSets/GDCdata/TCGA-LGG/Transcriptome_Profiling/Gene_Expression_Quantification")) {
  GDCdownload(query = query_rna)
}

data.rna <- GDCprepare(query_rna, summarizedExperiment = TRUE)
rm(query_rna)
```
"Genome of reference: hg38" in the output of the query, so I guess this is harmonized.




```{r}

info.rna <- as.matrix(rowData(data.rna))
assay.rna <- t(as.matrix(assay(data.rna)))

rm(data.rna)
```
Type of Data -> raw counts.



In "assay.rna", there are many types of RNA-seq, including "miRNA".
```{r}

unique(info.rna[, "gene_type"])
```



```{r}

assay.rna.coding <- get_assay_gene_type("protein_coding", assay.rna, info.rna)
assay.rna.mirna <- get_assay_gene_type("miRNA", assay.rna, info.rna)


dim(assay.rna)
dim(assay.rna.coding)
dim(assay.rna.mirna)
```

## Save Original Assays

```{r}
write.csv(assay.rna.mirna, "original_assays/assay_rna_mirna.csv", row.names = TRUE)
write.csv(assay.rna.coding, "original_assays/assay_rna_coding.csv", row.names = TRUE)
```



## Arrange the names (for later PCGSEA)

```{r}

colnames(assay.rna.coding) <- sub("\\.\\d+$", "", colnames(assay.rna.coding))
colnames(assay.rna.mirna) <- sub("\\.\\d+$", "", colnames(assay.rna.mirna))
```


## Data Quality Check and Cleaning

# Uniquely identifying 
```{r}

ids_duplicated(assay.rna.coding)
ids_duplicated(assay.rna.mirna)
```


```{r}

dim(assay.rna.coding)
assay.rna.coding <- del_repl_barcodes(assay.rna.coding)
dim(assay.rna.coding)

dim(assay.rna.mirna)
assay.rna.mirna <- del_repl_barcodes(assay.rna.mirna)
dim(assay.rna.mirna)
```




#  Handling Missing Values

```{r}

sum(is.na(assay.rna.coding)) / prod(dim(assay.rna.coding))
sum(is.na(assay.rna.mirna)) / prod(dim(assay.rna.mirna))
```
The assays do not have missing values. 



```{r}
sum(assay.rna.coding==0)/prod(dim(assay.rna.coding)) *100
sum(assay.rna.mirna==0)/prod(dim(assay.rna.mirna))*100
```

# Outliers


```{r}

target_assays <- list(t(assay.rna.coding), t(assay.rna.mirna))
```


Remove Low-Count Genes (retain genes that have counts > 1 in at least five samples)
```{r}

lapply(target_assays, dim)
for (assay_idx in 1:length(target_assays)){
  keep <- rowSums(cpm(target_assays[[assay_idx]]) > 1) >= 5
  target_assays[[assay_idx]] <- target_assays[[assay_idx]][keep,]
}
lapply(target_assays, dim)
```



## EDA

# Transformation


Normalizing according to the library size
```{r}

for (assay_idx in 1:length(target_assays)){
  dge_assay <- DGEList(counts = target_assays[[assay_idx]])
  target_assays[[assay_idx]] <- normLibSizes(dge_assay, method = "TMM")
  
}
```


Voom transformation to stabilize the variance and convert the counts into log2-CPM
```{r}

for (assay_idx in 1: length(target_assays)){
  temp <- voom(target_assays[[assay_idx]], plot = F)
  target_assays[[assay_idx]] <- as.matrix(temp$E)
}
```


```{r}
target_assays <- lapply(target_assays, t)
```



# Pre-feature selection

```{r}

sum(assay.rna.coding == 0) / prod(dim(assay.rna.coding))
sum(assay.rna.mirna == 0)/ prod(dim(assay.rna.mirna))
```

What are the intervals of the variance for each assay?
```{r}

for (assay_idx in 1:length(target_assays)){
  var_vec <- apply(target_assays[[assay_idx]], MARGIN = 2, function(x) var(x, na.rm = TRUE))
  print(summary(var_vec))
}

```


```{r}
var_mrna <- apply(target_assays[[1]], 2, var)
df_mrna <- data.frame(variance = var_mrna)

var_mirna <- apply(target_assays[[2]], 2, var)
df_mirna <- data.frame(variance = var_mirna)


text_size <- 2
plot1 <- ggplot(df_mrna, aes(y = variance)) +
    geom_boxplot(fill = "#FF7F32", color = "black", outlier.color = "#00B3B3") +  
  geom_hline(aes(yintercept = quantile(variance, 0.5)),  
             color = "black", lwd = 0.8, linetype = "dashed") +  
  labs(y = "Feature Variance", x = "") +  
  theme(
    axis.ticks.x = element_blank(),
    axis.text.x = element_blank(),
      axis.text.y = element_text(size=rel(text_size), hjust=1, color='black'),
     
      axis.title.y= element_text(size=rel(text_size), hjust=1, color='black'),
      axis.title.x=element_text(size=rel(text_size), vjust=1, color='black'),
      legend.text = element_text(size = rel(text_size), color = 'black'),
       legend.title = element_text(size = rel(text_size), color = 'black'),
     panel.grid.minor = element_blank(), 
      panel.border = element_rect(color = "black", fill = NA, size = 0.8),
      panel.background = element_blank(),
    )

plot2 <- ggplot(df_mirna, aes(y = variance)) +
   geom_boxplot(fill = "#FF7F32", color = "black", outlier.color = "#00B3B3") +  
  geom_hline(aes(yintercept = quantile(variance, 0.2)),  
             color = "black", lwd = 0.8, linetype = "dashed") +  
  labs(y = "Feature Variance", x = "") +  
  theme(
    axis.ticks.x = element_blank(),
    axis.text.x = element_blank(), 
      axis.text.y = element_text(size=rel(text_size), hjust=1, color='black'),
      axis.title.y= element_text(size=rel(text_size), hjust=1, color='black'),
      axis.title.x=element_text(size=rel(text_size), vjust=1, color='black'),
      legend.text = element_text(size = rel(text_size), color = 'black'),
       legend.title = element_text(size = rel(text_size), color = 'black'),
     panel.grid.minor = element_blank(), 
      panel.border = element_rect(color = "black", fill = NA, size = 0.8),
      panel.background = element_blank(),
    )


plot1
plot2
ggsave(filename = "mrna_feature_var.pdf", plot = plot1, width = 9, height = 7, dpi = 300)
ggsave(filename = "mirna_feature_var.pdf", plot = plot2, width = 9, height = 7, dpi = 300)
```





Keeping the 50% of the top most variable genes (for mRNA and lncRNA) and 80% of miRNA
```{r}


sorted_mrna <- order(var_mrna, decreasing = TRUE)
sorted_mirna <- order(var_mirna, decreasing = TRUE)


# keep the top 50% most variable genes for mRNA/lncRNA
top_50_mrna <- sorted_mrna[1: floor(length(var_mrna) * 0.5)]
target_assays[[1]] <- target_assays[[1]][,top_50_mrna ]


# keep the top 80% most variable genes for miRNA
top_80_mirna <- sorted_mirna[1: floor(length(var_mirna) * 0.8)]
target_assays[[2]] <- target_assays[[2]][,top_80_mirna ]

```






# Normalization 
```{r}

target_assays <- lapply(target_assays, function(assay) scale(assay, center = TRUE, scale = TRUE))
```

```{r}

assay.rna.coding <- target_assays[[1]]
assay.rna.mirna <- target_assays[[2]]
```



## Arrange the info matrices

```{r}
rownames(info.rna) <- sub("\\.\\d+$", "", rownames(info.rna))

info.rna.mirna <- info.rna[colnames(assay.rna.mirna), , drop = FALSE]
info.rna.coding <- info.rna[colnames(assay.rna.coding), , drop = FALSE]
```




## Save Final Matrices
```{r}

write.csv(assay.rna.mirna, "DataSets/processed_assays/assay_rna_mirna.csv", row.names = TRUE)
write.csv(assay.rna.coding, "DataSets/processed_assays/assay_rna_coding.csv", row.names = TRUE)

write.csv(info.rna.mirna, "DataSets/processed_assays/info_rna_mirna.csv", row.names = TRUE)
write.csv(info.rna.coding, "DataSets/processed_assays/info_rna_coding.csv", row.names = TRUE)
```

